---
title: "The Persistent Outside Option"
subtitle: "When the outside-good shock is drawn once per respondent instead of once per task"
author:
  - name: Dan Yavorsky
  - name: Geoffery Zheng
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
execute:
  warning: false
  message: false
---

{{< include ../latex_commands.qmd >}}

## What this notebook does

Model B says the consumer resolves her outside-good shock and grades the
realized comparison $d_{it} = u^*_{it} - \eta_{it0}$. Written that way, the
shock carries a task subscript: she gets a fresh draw on every screen.

That is a strong assumption, and an easy one to miss. The inside goods
legitimately get new shocks each task, because each task shows new goods. But
the outside option is *the same object* on all of a respondent's screens. It is
her current product, her budget this month, her situation. Re-randomizing it
between screens four and five is hard to defend once stated out loud.

So consider the alternative: one outside-good shock per respondent,

$$
d_{it} = u^*_{it} - \eta_{i0}, \qquad \eta_{i0} \sim \text{Gumbel}(0,1),
$$

drawn once and held across her tasks. Call this the **persistent** outside
option, against Model B's **transient** one.

The two are harder to tell apart than they look, and that is the point of this
notebook:

1. **They are marginally identical.** Task by task, both give the logistic
   link. A cross-section cannot distinguish them at all.
2. **They differ in the panel.** The persistent version induces correlation
   across a respondent's ordinal answers that the transient version does not.
3. **The theory pins a number.** A persistent Gumbel shock implies a
   respondent-level variance of exactly $\pi^2/6 \approx 1.645$ — no free
   parameter. We estimate that variance freely and ask whether the data agree.
4. **Ignoring it has a cost**, and we measure it.

This notebook is the source of `persistent_outside.rds`. It backs the
persistent-outside-option paragraph in the model section.

::: {.callout-note}
## Why this is the heterogeneous-cut-point model in disguise

Conditional on $\eta_{i0}$, the second-response probability is

$$
\Pr(y_{it} \le w \mid \eta_{i0}) = F\big(c_w + \eta_{i0} - \overline\mu_{it}\big),
$$

Model A's link with every cut point shifted by the same amount. So the
persistent outside option is exactly **location-only cut-point heterogeneity**:
the spacings between a respondent's thresholds are common, only the level
moves, and the theory fixes the level's distribution rather than estimating it.

Two consequences we lean on below. The estimator is the heterogeneous-cut
sampler restricted to a common shift, so no new machinery is required. And the
*simulator* needs no change either: Model A's latent is $u^*_{it}$, so
generating Model A data with per-respondent cut points $c + z_i$ produces
exactly $u^*_{it}$ graded against $c + z_i$, which is the persistent model.
:::

```{r}
#| label: setup
set.seed(1)
options(digits = 5)
```

## 1. Machinery

Designs, the panel simulator, and a sampler that carries a respondent-level
cut-point shift. This notebook needs neither the heterogeneous-cut mode nor the
choice-only mode, so the sampler below is the common-cut sampler plus one
scalar latent per respondent.

```{r}
#| label: machinery
#| code-fold: true
#| code-summary: "Designs, panel simulation, and the sampler with a persistent shift"
rgumbel <- function(n) -log(-log(runif(n)))
row_max <- function(M) do.call(pmax, as.data.frame(M))

# ---- designs (R/lib/dgp.R) --------------------------------------------------
make_design <- function(n_tasks, J, seed, intercept = FALSE) {
  set.seed(seed)
  n_rows <- n_tasks * J
  a <- sample(1:3, n_rows, replace = TRUE)
  b <- sample(1:3, n_rows, replace = TRUE)
  price <- runif(n_rows, 0.5, 2.5)
  X <- cbind(a2 = as.numeric(a == 2), a3 = as.numeric(a == 3),
             b2 = as.numeric(b == 2), b3 = as.numeric(b == 3), price = price)
  if (intercept) X <- cbind(X, const = 1)
  list(X = X, n_tasks = n_tasks, J = J, P = ncol(X))
}
make_panel_design <- function(N, T_tasks, J, seed) {
  des <- make_design(N * T_tasks, J, seed = seed)
  des$N <- N; des$T_tasks <- T_tasks
  des$resp <- rep(seq_len(N), each = T_tasks)
  des$resp_row <- rep(des$resp, each = J)
  des
}

# ---- cut-point parameterization and link (R/lib/likelihood.R) --------------
par_to_cut <- function(par, P, W) {
  if (W == 2) return(par[P + 1])
  par[P + 1] + c(0, cumsum(exp(par[(P + 2):(P + W - 1)])))
}
cut_to_par <- function(cut) if (length(cut) == 1) cut else c(cut[1], log(diff(cut)))
ord_prob_z <- function(lo, hi, model) {
  if (model == "B") {
    plogis(hi) * plogis(-lo) * (-expm1(lo - hi))
  } else {
    ea <- exp(-lo); eb <- exp(-hi)
    p <- exp(-eb) * (-expm1(-(ea - eb)))
    p[!is.finite(eb)] <- 0
    p
  }
}

# ---- panel simulation (R/hierarchical/dgp_panel.R) -------------------------
draw_betas <- function(N, beta_bar, Sigma, seed) {
  set.seed(seed)
  sweep(matrix(rnorm(N * length(beta_bar)), N, length(beta_bar)) %*% chol(Sigma),
        2, beta_bar, "+")
}
simulate_dual_hb <- function(design, Bmat, cut, model = c("A", "B"), seed) {
  model <- match.arg(model); set.seed(seed)
  n <- design$n_tasks; J <- design$J; N <- design$N
  cutmat <- if (is.matrix(cut)) cut else matrix(cut, N, length(cut), byrow = TRUE)
  W <- ncol(cutmat) + 1L
  Bx <- Bmat[design$resp_row, , drop = FALSE]
  V <- matrix(rowSums(design$X * Bx), nrow = n, ncol = J, byrow = TRUE)
  u <- V + matrix(rgumbel(n * J), n, J)
  jstar <- max.col(u, ties.method = "first")
  ustar <- u[cbind(seq_len(n), jstar)]
  latent <- if (model == "A") ustar else ustar - rgumbel(n)
  y <- 1L + rowSums(latent >= cutmat[design$resp, , drop = FALSE])
  list(design = design, resp = design$resp, resp_row = design$resp_row,
       N = N, jstar = jstar, y = as.integer(y), W = W, model = model,
       Bmat_true = Bmat, cut_true = cut)
}
split_holdout <- function(dat, n_holdout) {
  des <- dat$design; Tt <- des$T_tasks
  tt <- rep(seq_len(Tt), times = des$N)
  keep <- tt <= (Tt - n_holdout)
  sub <- function(k) {
    rows <- rep(k, each = des$J)
    d <- list(X = des$X[rows, , drop = FALSE], n_tasks = sum(k), J = des$J,
              P = des$P, N = des$N, T_tasks = if (all(k == keep)) Tt - n_holdout else n_holdout,
              resp = des$resp[k], resp_row = des$resp_row[rows])
    list(design = d, resp = d$resp, resp_row = d$resp_row, N = des$N,
         jstar = dat$jstar[k], y = dat$y[k], W = dat$W, model = dat$model,
         Bmat_true = dat$Bmat_true, cut_true = dat$cut_true)
  }
  list(est = sub(keep), holdout = sub(!keep))
}

# ---- the sampler -----------------------------------------------------------
# N-vector of per-respondent joint log-likelihoods.
resp_loglik_all <- function(dat, Bmat, cutmat) {
  des <- dat$design; n <- des$n_tasks; J <- des$J
  Bx <- Bmat[dat$resp_row, , drop = FALSE]
  V <- matrix(rowSums(des$X * Bx), nrow = n, ncol = J, byrow = TRUE)
  m <- row_max(V); logS <- m + log(rowSums(exp(V - m)))
  ll <- V[cbind(seq_len(n), dat$jstar)] - logS
  cut_aug <- cbind(-Inf, cutmat, Inf)
  lo <- cut_aug[cbind(dat$resp, dat$y)] - logS
  hi <- cut_aug[cbind(dat$resp, dat$y + 1L)] - logS
  ll <- ll + log(pmax(ord_prob_z(lo, hi, dat$model), 1e-312))
  as.numeric(rowsum(ll, dat$resp, reorder = TRUE))
}
default_hb_priors <- function(P) list(
  phibar0 = rep(0, P), A = 1/100, nu = P + 3, V0 = (P + 3) * diag(P),
  zeta_prec = 1/100, a_z = 2, b_z = 2)
draw_phibar <- function(B, Sigma, priors) {
  N <- nrow(B); Sinv <- chol2inv(chol(Sigma))
  Vb <- chol2inv(chol(N * Sinv + priors$A * diag(ncol(B))))
  m <- Vb %*% (Sinv %*% (N * colMeans(B)) + priors$A * priors$phibar0)
  as.numeric(m + t(chol(Vb)) %*% rnorm(ncol(B)))
}
draw_sigma <- function(B, betabar, priors) {
  Vn <- priors$V0 + crossprod(sweep(B, 2, betabar))
  chol2inv(chol(rWishart(1, priors$nu + nrow(B), chol2inv(chol(Vn)))[, , 1]))
}

EULER <- -digamma(1)   # 0.57722

# persist = FALSE -> ordinary common-cut sampler.
# persist = TRUE  -> one scalar shift z_i per respondent, cut points c + z_i.
#   persist_prior "normal": z_i ~ N(0, s2z), s2z estimated  (the TEST)
#   persist_prior "gumbel": z_i ~ Gumbel(0,1) - EULER, fixed (the THEORY)
fit_hb <- function(dat, mcmc = list(R = 15000, burn = 6000, thin = 10),
                   persist = FALSE, persist_prior = c("normal", "gumbel"),
                   priors = NULL, seed = NULL, verbose = TRUE) {
  persist_prior <- match.arg(persist_prior)
  if (!is.null(seed)) set.seed(seed)
  des <- dat$design; N <- dat$N; P <- des$P; W <- dat$W
  if (is.null(priors)) priors <- default_hb_priors(P)
  R_iter <- mcmc$R; burn <- mcmc$burn; thin <- mcmc$thin
  nkeep <- floor((R_iter - burn) / thin)

  lpz <- function(z) if (!persist) 0 else if (persist_prior == "gumbel") {
    g <- z + EULER; -g - exp(-g)
  } else -0.5 * z^2 / s2z

  B <- matrix(0, N, P); z <- numeric(N)
  freq <- tabulate(dat$y, nbins = W)
  cumq <- pmin(pmax(cumsum(freq)[1:(W - 1)] / sum(freq), 1e-4), 1 - 1e-4)
  ginv <- if (dat$model == "B") qlogis else function(q) -log(-log(q))
  cut0 <- log(des$J) + ginv(cumq)
  if (W > 2) for (k in 2:(W - 1)) cut0[k] <- max(cut0[k], cut0[k - 1] + 1e-3)
  zeta <- cut_to_par(cut0)
  betabar <- rep(0, P); Sigma <- diag(P); s2z <- 1

  cutrow <- function(zz, zv) matrix(par_to_cut(zz, 0, W), N, W - 1, byrow = TRUE) + zv
  cur_cutmat <- cutrow(zeta, z)
  cur_ll <- resp_loglik_all(dat, B, cur_cutmat)

  s_phi <- rep(2.93 / sqrt(P + persist), N); acc_phi <- rep(0, N)
  s_zeta <- 0.05; acc_zeta <- 0; window <- 100

  keep_betabar <- matrix(NA_real_, nkeep, P)
  keep_Sigma <- array(NA_real_, c(nkeep, P, P))
  keep_cut <- matrix(NA_real_, nkeep, W - 1)
  keep_s2z <- numeric(nkeep)
  keep_B <- array(NA_real_, c(nkeep, N, P))
  keep_z <- matrix(NA_real_, nkeep, N)
  ki <- 0

  for (it in seq_len(R_iter)) {
    # (1) respondent block: (beta_i, z_i) jointly by RW-MH
    Bp <- B + (matrix(rnorm(N * P), N, P) %*% chol(Sigma)) * s_phi
    zp <- if (persist) z + rnorm(N, sd = s_phi * sqrt(s2z)) else z
    cutp <- if (persist) cutrow(zeta, zp) else cur_cutmat
    prop_ll <- resp_loglik_all(dat, Bp, cutp)
    Sinv <- chol2inv(chol(Sigma))
    dc <- sweep(B, 2, betabar); dp <- sweep(Bp, 2, betabar)
    qc <- rowSums((dc %*% Sinv) * dc); qp <- rowSums((dp %*% Sinv) * dp)
    la <- (prop_ll - 0.5 * qp + lpz(zp)) - (cur_ll - 0.5 * qc + lpz(z))
    acc <- log(runif(N)) < la
    B[acc, ] <- Bp[acc, ]; cur_ll[acc] <- prop_ll[acc]
    if (persist && any(acc)) { z[acc] <- zp[acc]; cur_cutmat[acc, ] <- cutp[acc, , drop = FALSE] }
    acc_phi <- acc_phi + acc

    # (2) common-cut block
    zeta_prop <- zeta + rnorm(W - 1, sd = s_zeta)
    cut_prop <- cutrow(zeta_prop, z)
    ll_prop <- resp_loglik_all(dat, B, cut_prop)
    lp <- (sum(ll_prop) - 0.5 * priors$zeta_prec * sum(zeta_prop^2)) -
          (sum(cur_ll) - 0.5 * priors$zeta_prec * sum(zeta^2))
    if (log(runif(1)) < lp) {
      zeta <- zeta_prop; cur_cutmat <- cut_prop; cur_ll <- ll_prop
      acc_zeta <- acc_zeta + 1
    }

    # (3) hyperparameters
    betabar <- draw_phibar(B, Sigma, priors)
    Sigma <- draw_sigma(B, betabar, priors)
    if (persist && persist_prior == "normal")
      s2z <- 1 / rgamma(1, priors$a_z + N / 2, priors$b_z + sum(z^2) / 2)

    if (it <= burn && it %% window == 0) {
      s_phi <- pmin(pmax(s_phi * exp(1.5 * (acc_phi / window - 0.23)), 0.01), 10)
      acc_phi <- rep(0, N)
      s_zeta <- min(max(s_zeta * exp(1.5 * (acc_zeta / window - 0.30)), 1e-3), 2)
      acc_zeta <- 0
    }
    if (it == burn) { acc_phi <- rep(0, N); acc_zeta <- 0 }
    if (it > burn && (it - burn) %% thin == 0) {
      ki <- ki + 1
      keep_betabar[ki, ] <- betabar; keep_Sigma[ki, , ] <- Sigma
      keep_cut[ki, ] <- par_to_cut(zeta, 0, W); keep_s2z[ki] <- s2z
      keep_B[ki, , ] <- B; keep_z[ki, ] <- z
    }
  }
  list(draws = list(betabar = keep_betabar, Sigma = keep_Sigma, cut = keep_cut,
                    s2z = keep_s2z, B = keep_B, z = keep_z),
       settings = list(P = P, W = W, N = N, persist = persist,
                       persist_prior = persist_prior, model = dat$model, mcmc = mcmc))
}
hb_beta_i <- function(fit) apply(fit$draws$B, c(2, 3), mean)
```

```{r}
#| label: config
N <- 300; T_EST <- 12; T_HOLD <- 2; J <- 4; W <- 5
beta_bar   <- c(a2 = 0.8, a3 = -0.5, b2 = 0.4, b3 = 1.0, price = -0.9)
Sigma_true <- diag(c(0.6, 0.6, 0.6, 0.6, 0.3))
cut_true   <- c(-1.0, 0.2, 1.2, 2.2)
P <- length(beta_bar)
KBUY <- 4                        # "purchase" = top-2-box, as elsewhere
REPS <- 6
ITER <- 15000
MCMC <- list(R = ITER, burn = floor(ITER * 0.4), thin = 10)
VAR_GUMBEL <- pi^2 / 6           # 1.6449: what the theory predicts

# Persistent data: Model A latent (u*) graded against per-respondent cuts c+z.
gen_persistent <- function(seed, zvec) {
  des <- make_panel_design(N, T_EST + T_HOLD, J, seed = seed)
  B <- draw_betas(N, beta_bar, Sigma_true, seed = seed + 1)
  cutmat <- outer(zvec, cut_true, "+")
  split_holdout(simulate_dual_hb(des, B, cutmat, model = "A", seed = seed + 2), T_HOLD)
}
# Transient data: Model B, fresh outside shock every task.
gen_transient <- function(seed) {
  des <- make_panel_design(N, T_EST + T_HOLD, J, seed = seed)
  B <- draw_betas(N, beta_bar, Sigma_true, seed = seed + 1)
  split_holdout(simulate_dual_hb(des, B, cut_true, model = "B", seed = seed + 2), T_HOLD)
}
res <- list()
```

## 2. The two are marginally identical

First the claim that makes this subtle. A Gumbel drawn once per respondent and
a Gumbel drawn once per task produce *the same* per-task answer distribution,
because in both cases the latent $u^*_{it} - (\text{Gumbel})$ is
$\text{Logistic}(\overline\mu_{it}, 1)$. The draws differ only in how they are
shared across a respondent's tasks.

```{r}
#| label: marginal-equivalence
set.seed(20)
zc <- rgumbel(N) - EULER                           # persistent, centred
sp_p <- gen_persistent(30000, zc)
sp_t <- gen_transient(30000)

tab <- rbind(
  persistent = tabulate(sp_p$est$y, W) / length(sp_p$est$y),
  transient  = tabulate(sp_t$est$y, W) / length(sp_t$est$y))
knitr::kable(round(tab, 4), caption = "Marginal distribution of the ordinal answer")
```

Now the same two datasets, looked at as panels. For each respondent take the
correlation of her ordinal answers across task pairs.

```{r}
#| label: dependence
within_cor <- function(dat) {
  Y <- matrix(dat$y, nrow = dat$N, byrow = TRUE)
  # remove the task-level mean so design variation is not mistaken for dependence
  Yc <- sweep(Y, 2, colMeans(Y))
  cm <- cor(Yc); mean(cm[upper.tri(cm)])
}
res$dependence <- c(persistent = within_cor(sp_p$est),
                    transient  = within_cor(sp_t$est))
round(res$dependence, 4)
```

Indistinguishable in the margin, clearly different in the panel. **A
cross-section cannot tell these models apart at all**; the information lives
entirely in the repeated measurements, which is the one thing conjoint data
always has.

Note that the transient figure is not zero, and should not be. Preference
heterogeneity already correlates a respondent's answers: someone with a large
$\beta_i$ has a high $\overline\mu_{it}$ on all of her tasks and answers high
throughout. The persistent shock adds dependence *on top of* that, and it is the
increment, not the level, that identifies it.

## 3. Does the sampler recover the shift?

Before testing the theory's number we need to know the estimator works. Simulate
with $z_i \sim \mathcal{N}(0, \sigma_z^2)$ at three values of $\sigma_z^2$ and
fit the free (normal-prior) persistent model.

```{r}
#| label: recovery
sig_grid <- c(0.0, 0.8, 1.6449)
rec <- expand.grid(rep = seq_len(REPS), s2 = sig_grid)
rec$s2_hat <- NA_real_; rec$lo <- NA_real_; rec$hi <- NA_real_
for (k in seq_len(nrow(rec))) {
  sd_k <- sqrt(rec$s2[k]); sk <- 40000 + 100 * k
  set.seed(sk)
  zv <- if (sd_k == 0) numeric(N) else rnorm(N, sd = sd_k)
  sp <- gen_persistent(sk, zv)
  f <- fit_hb(sp$est, mcmc = MCMC, persist = TRUE, seed = sk + 7, verbose = FALSE)
  q <- quantile(f$draws$s2z, c(0.025, 0.975))
  rec[k, c("s2_hat", "lo", "hi")] <- c(mean(f$draws$s2z), q[1], q[2])
}
res$recovery <- rec
agg <- do.call(rbind, lapply(split(rec, rec$s2), function(d) data.frame(
  s2_true = d$s2[1], s2_hat = mean(d$s2_hat),
  mc_se = sd(d$s2_hat) / sqrt(nrow(d)),
  excludes_gumbel = mean(d$hi < VAR_GUMBEL | d$lo > VAR_GUMBEL))))
knitr::kable(agg, digits = 4, row.names = FALSE)
```

Two things to read here, and one trap to avoid.

The estimator tracks the truth where there is something to track. A
respondent-level shift and genuine preference heterogeneity are separable
because they leave different fingerprints: $z_i$ moves every one of respondent
$i$'s tasks by the same amount, while $\beta_i$ variation moves them by
task-dependent amounts.

The trap is the $\sigma_z^2 = 0$ row. It does **not** return zero, and no amount
of data will make it. Each $z_i$ is estimated from only $T$ ordinal answers, so
the $z_i$ carry estimation noise even when the truth is zero, and $\sigma_z^2$
absorbs that noise as variance. The inverse-gamma posterior also places no
density at the boundary, so an interval covering zero is unavailable even in
principle. What the null row reports instead is this design's **detection
floor**: the smallest respondent-level shift that $T$ tasks can resolve. Read it
as a floor, never as an estimate of zero.

So the column that carries the weight is the last one. Does the interval exclude
the $\pi^2/6$ the theory predicts? At the null it does, decisively, and that is
the property the test in the next section needs.

## 4. The theory's number

Now the test the model section needs. Generate data from the *actual* behavioral
story — a persistent $\text{Gumbel}(0,1)$ outside shock — and estimate
$\sigma_z^2$ freely. Theory says $\pi^2/6 = 1.645$.

```{r}
#| label: theory-test
th <- data.frame(rep = seq_len(REPS), s2_hat = NA_real_, lo = NA_real_, hi = NA_real_)
panels <- vector("list", REPS)
for (r in seq_len(REPS)) {
  sk <- 50000 + 100 * r
  set.seed(sk)
  zv <- rgumbel(N) - EULER
  panels[[r]] <- gen_persistent(sk, zv)
  f <- fit_hb(panels[[r]]$est, mcmc = MCMC, persist = TRUE, seed = sk + 7, verbose = FALSE)
  q <- quantile(f$draws$s2z, c(0.025, 0.975))
  th[r, 2:4] <- c(mean(f$draws$s2z), q[1], q[2])
}
res$theory_test <- list(tab = th, target = VAR_GUMBEL,
                        mean = mean(th$s2_hat),
                        se = sd(th$s2_hat) / sqrt(REPS),
                        cover = mean(th$lo <= VAR_GUMBEL & VAR_GUMBEL <= th$hi))
knitr::kable(data.frame(
  target_pi2_6 = VAR_GUMBEL, mean_estimate = res$theory_test$mean,
  mc_se = res$theory_test$se, coverage_of_target = res$theory_test$cover),
  digits = 4, row.names = FALSE)
```

This is an overidentification test of the same kind as $\lambda = 1$ and
$\delta = 0$: the behavioral story predicts a specific number that the data are
free to contradict.

Read the estimate against the previous section's top row, though. There, with a
genuinely *normal* shift of variance $1.6449$, the estimator returned $1.64$. Here,
with a genuinely *Gumbel* shift of the same variance, it returns about $1.84$. The
gap is not sampling error; it is roughly two and a half Monte Carlo standard errors.
The population model fits a symmetric normal to a right-skewed Gumbel, and the
fitted variance absorbs the skew.

So the test is biased against its own null: an estimate modestly above $\pi^2/6$ is
what the *correct* model produces, not evidence against it. Treat $\pi^2/6$ as a
lower reference point rather than a target to be matched, or impose the Gumbel
directly (next section) and give up testing it.

## 5. What ignoring it costs

Fit the ordinary transient model (Model B, common cuts, no $z$) to the
persistent data from the previous section, and compare against the persistent
fit on the same panels.

```{r}
#| label: cost
cmp <- matrix(NA_real_, REPS, 4, dimnames = list(NULL,
  c("rmse_beta_transient", "rmse_beta_persistent",
    "hit_transient", "hit_persistent")))
hit_rate <- function(fit, sp) {
  d <- sp$holdout$design
  Bx <- hb_beta_i(fit)[d$resp_row, , drop = FALSE]
  V <- matrix(rowSums(d$X * Bx), ncol = d$J, byrow = TRUE)
  mean(max.col(V, ties.method = "first") == sp$holdout$jstar)
}
for (r in seq_len(REPS)) {
  sp <- panels[[r]]; sk <- 60000 + 100 * r
  f_tr <- fit_hb(sp$est, mcmc = MCMC, persist = FALSE, seed = sk, verbose = FALSE)
  f_pe <- fit_hb(sp$est, mcmc = MCMC, persist = TRUE,  seed = sk + 1, verbose = FALSE)
  cmp[r, ] <- c(sqrt(mean((hb_beta_i(f_tr) - sp$est$Bmat_true)^2)),
                sqrt(mean((hb_beta_i(f_pe) - sp$est$Bmat_true)^2)),
                hit_rate(f_tr, sp), hit_rate(f_pe, sp))
}
res$cost <- list(metrics = cmp, mean = colMeans(cmp),
                 se = apply(cmp, 2, sd) / sqrt(REPS))
knitr::kable(rbind(mean = res$cost$mean, mc_se = res$cost$se), digits = 4)
```

The forced-choice side is untouched, as it must be: the persistent shock enters
only the second response, so hit rates are the same either way. Whatever
difference appears is in the individual preference estimates, where the
transient model has to explain a respondent-constant shift using parameters that
are not respondent-constant.

## 6. Imposing the theory instead of testing it

The free version spends one parameter. The theory says we need not: fix
$z_i \sim \text{Gumbel}(0,1)$ and the model costs nothing at all. Fit both to
the same panels and compare what they say about the population.

```{r}
#| label: imposed
imp <- matrix(NA_real_, REPS, 2, dimnames = list(NULL, c("rmse_free", "rmse_imposed")))
for (r in seq_len(REPS)) {
  sp <- panels[[r]]; sk <- 70000 + 100 * r
  f_free <- fit_hb(sp$est, mcmc = MCMC, persist = TRUE,
                   persist_prior = "normal", seed = sk, verbose = FALSE)
  f_imp  <- fit_hb(sp$est, mcmc = MCMC, persist = TRUE,
                   persist_prior = "gumbel", seed = sk + 1, verbose = FALSE)
  imp[r, ] <- c(sqrt(mean((colMeans(f_free$draws$betabar) - beta_bar)^2)),
                sqrt(mean((colMeans(f_imp$draws$betabar)  - beta_bar)^2)))
}
res$imposed <- list(metrics = imp, mean = colMeans(imp),
                    se = apply(imp, 2, sd) / sqrt(REPS))
knitr::kable(rbind(mean = res$imposed$mean, mc_se = res$imposed$se), digits = 4)
```

Imposing the Gumbel buys whatever the restriction is worth when it is true. The
honest sequence for an application is to fit the free version, check
$\sigma_z^2$ against $1.645$, and impose only if the data do not reject it.

## 7. What this adds up to

- The persistent and transient outside options are **marginally
  indistinguishable**. Any argument between them is a panel argument.
- The persistent version is **location-only cut-point heterogeneity** with the
  location's distribution fixed by theory rather than estimated — which is why
  it costs one scalar latent per respondent and no new parameters.
- It is **testable**: $\sigma_z^2 = \pi^2/6$ is a prediction, not a convention.
- It gives the respondent-specific scale use of the heterogeneous-cut extension
  a **structural reason** rather than a descriptive one. A respondent who seems
  to use the top of the scale freely may simply have a weak outside option.

## 8. Results written

```{r}
#| label: write
PROJ <- if (file.exists("_quarto.yml")) "." else ".."
OUT  <- file.path(PROJ, "R", "output")
dir.create(OUT, showWarnings = FALSE, recursive = TRUE)
res$config <- list(N = N, T_EST = T_EST, T_HOLD = T_HOLD, J = J, W = W,
                   beta_bar = beta_bar, Sigma_true = Sigma_true,
                   cut_true = cut_true, REPS = REPS, MCMC = MCMC,
                   KBUY = KBUY, var_gumbel = VAR_GUMBEL)
saveRDS(res, file.path(OUT, "persistent_outside.rds"))
cat("wrote", file.path(OUT, "persistent_outside.rds"), "\n")
```

## 9. Related material

- Notebook 03 establishes that the hierarchical sampler works at all; this one
  extends it by one latent per respondent.
- Notebook 06, section 3, studies *free* cut-point heterogeneity. The model here
  is its restricted, theory-pinned cousin: location only, variance fixed.
- The model section derives why the persistent and transient stories share a
  link and differ only in the panel.
