---
title: "How the Ordinal Dual-Response Models Are Identified"
subtitle: "A plain-language walkthrough of the aggregate model, the hierarchical model with common cut points, and the hierarchical model with respondent-specific cut points"
author: "Dan Yavorsky"
date: 2026-09-15
bibliography: ../references.bib
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: false
execute:
  warning: false
  message: false
---

::: {.callout-note}
## Purpose and audience

This note explains, in as plain a language as the subject allows, *why* each of our
three models can be estimated at all, and what has to be true of the experimental
design for that to work. It assumes you know what a regression is and that you have
seen a logit once. It does not assume you have seen a multinomial logit, an ordered
logit, or a hierarchical Bayesian model.

Everything here is a restatement of Proposition 1 in Section 2 of the paper and the
remarks that follow it. Nothing here is new. The point is to make the argument
followable without a proof.

Not part of the manuscript. Written for teaching and for discussion.
:::

## What "identification" means

Before any model, the idea.

A statistical model is a machine that turns parameters into predicted probabilities.
You feed in a candidate set of numbers, and it tells you how likely each possible
survey answer is. Estimation runs the machine backwards: you have the answers, and
you want the numbers.

**A parameter is identified if no other parameter value predicts exactly the same
thing.** That is the whole definition. If two different parameter vectors
$\theta$ and $\theta'$ produce *identical* probabilities for every answer a
respondent could give, on every task you could possibly show them, then no amount of
data will ever separate the two. Not a bigger sample, not a better optimizer, not a
smarter prior. The data only ever see predictions, and the predictions are the same.

So the way to check identification is not to stare at the estimator. It is to ask:

> Can I find two different parameter values that make exactly the same predictions?

If you can, the model is unidentified and you have just found the direction in which
it fails. If you can prove you cannot, the model is identified. That is exactly how
the proof of Proposition 1 is organized, and it is how this note is organized.

A related idea will matter in Section 3. A parameter can be *technically* identified
but *practically* hopeless: two parameter values that predict almost, but not quite,
the same thing. Then the data can separate them in principle, but only with far more
observations than you have. Statisticians call this weak identification. It is the
normal condition of individual-level parameters in conjoint analysis, and it is the
reason hierarchical models exist.

## A 30-second recap of the model

Each respondent sees $J$ "inside" goods and answers two questions.

**Question 1, the forced choice.** Which of these $J$ do you prefer? She picks the
one with the highest utility $u_j = \mathbf{x}_j'\beta + \eta_j$, where $\mathbf{x}_j$
is the good's attributes, $\beta$ is her taste for those attributes, and $\eta_j$ is
a random shock. The outside good, "buy nothing," is *not* on the screen.

**Question 2, the ordinal report.** How likely are you to actually buy the one you
picked? She answers on a $W$-point scale, "very unlikely" through "very likely."

Our model says the second answer is about the *level* of the utility she just
achieved, $u^* = \max_j u_j$, compared against a set of thresholds
$c_1 < c_2 < \cdots < c_{W-1}$ that define the scale labels. From the researcher's
side, $u^*$ has a known distribution centered at

$$
\overline{\mu}_t \;=\; \ln \sum_{j=1}^{J} e^{\mathbf{x}_{tj}'\beta},
$$

the **inclusive value** of task $t$: one number summarizing how attractive the whole
screen was. The two answers then have probabilities

$$
\underbrace{\Pr(\text{picks } j) = \frac{e^{\mathbf{x}_{tj}'\beta}}{\sum_k e^{\mathbf{x}_{tk}'\beta}}}_{\text{forced choice}}
\qquad\qquad
\underbrace{\Pr(y_t \le w) = G\big(c_w - \overline{\mu}_t\big)}_{\text{ordinal report}}
$$

where $G$ is a link function (Gumbel under Model A, logistic under Model B; which one
does not matter for anything in this note).

The parameters to be found are $\beta$ and the cut points $\mathbf{c}$.

::: {.callout-tip}
## The one piece of algebra that drives everything

Suppose you add the same constant $\delta$ to the utility of **every** good on the
screen. What happens to the inclusive value?

$$
\ln \sum_j e^{V_j + \delta}
\;=\; \ln \left( e^{\delta} \sum_j e^{V_j} \right)
\;=\; \delta + \ln \sum_j e^{V_j}
\;=\; \delta + \overline{\mu}.
$$

**A common shift passes straight through the log-sum-exp.** Shift every alternative
by $\delta$ and the inclusive value moves by exactly $\delta$.

That one line is the source of every identification problem in this paper. Hold onto
it.
:::

```{r setup}
#| code-fold: true
#| code-summary: "Setup: load the paper's own likelihood code"
PROJ <- if (file.exists("_quarto.yml")) "." else ".."
source(file.path(PROJ, "R", "lib", "dgp.R"))
source(file.path(PROJ, "R", "lib", "likelihood.R"))
```

# Section 1. The aggregate model

The aggregate model assumes everyone shares one taste vector $\beta$ and one set of
cut points $\mathbf{c}$. This is the simplest case and it contains the entire
argument; Sections 2 and 3 are variations on it.

## 1.1 The division of labor between the two answers

The two survey questions tell you genuinely different things, and the cleanest way to
understand identification here is to keep them separate.

> **The forced choice sees only *differences*. The ordinal report sees *levels*.**

Both halves of that sentence need explaining.

### The forced choice sees only differences

When a respondent picks between goods, all that matters is which utility is biggest.
Adding the same amount to all of them changes nothing about which is biggest. This is
the oldest fact in discrete choice [@Train_2009, ch. 2]: *only differences in utility
matter*.

Concretely: suppose the screen shows three goods with utilities $1.2$, $-0.4$, $0.8$.
Now suppose we add $0.7$ to all three, giving $1.9$, $0.3$, $1.5$. The respondent's
ranking is unchanged, and so are the choice probabilities.

```{r d1}
V <- c(1.2, -0.4, 0.8)
softmax <- function(v) exp(v) / sum(exp(v))

rbind(original = softmax(V),
      shifted  = softmax(V + 0.7))
```

Identical. The forced choice is *blind* to any shift that moves everything together.

### The ordinal report sees levels

The second answer is different. It compares $\overline{\mu}_t$, a level, against the
cut points, which are also levels. If the whole screen gets better, $\overline{\mu}_t$
rises, it clears more thresholds, and answers shift toward "very likely." This is
precisely the information the forced choice throws away, and it is the reason the
dual-response design is worth running at all.

```{r d1b}
mubar <- function(v) log(sum(exp(v)))
cuts  <- c(-0.5, 0.7, 1.9)

ord_dist <- function(v) {
  m <- mubar(v)
  diff(c(0, plogis(cuts - m), 1)) |> round(3) |> rev()
}

rbind(`unattractive screen` = ord_dist(c(-1, -1.5, -0.8)),
      `attractive screen`   = ord_dist(c(1.2, -0.4, 0.8))) |>
  `colnames<-`(paste0("P(y=", 4:1, ")"))
```

The attractive screen pushes mass toward the top of the scale. The ordinal answers
carry information about levels that the forced choice cannot.

So: the forced choice pins down the *contrast* directions of $\beta$ (does she prefer
red to blue? is she price sensitive?), and the ordinal answers pin down *where the
whole thing sits* relative to the scale labels.

## 1.2 The gap: what slips past both questions

Now the problem. What if there is a direction in $\beta$ that moves *every* good on
*every* screen by the same amount?

Call that direction $v$, and suppose $\mathbf{x}_{tj}'v = \delta$ for every good $j$
and every task $t$. Then:

- The forced choice does not notice, by §1.1. Everything shifted together.
- The inclusive value moves to $\overline{\mu}_t + \delta$ in every task, by the
  log-sum-exp identity in the callout above.
- So if we *also* move every cut point to $c_w + \delta$, then
  $c_w + \delta - (\overline{\mu}_t + \delta) = c_w - \overline{\mu}_t$, and the
  ordinal probabilities do not move either.

Both questions give identical answers under $(\beta, \mathbf{c})$ and under
$(\beta + v, \mathbf{c} + \delta)$. The model is unidentified in the direction $v$.

That is the entire content of Proposition 1, which states it as a rank condition:

::: {.callout-important}
## The identification condition

$(\beta, \mathbf{c})$ is identified **if and only if** the only way to have
$\mathbf{x}_{tj}'v = \delta$ for every good and every task is $v = \mathbf{0}$ and
$\delta = 0$.

Equivalently: the design matrix $X$, with a column of ones stapled on, must have full
column rank.

In words: **no combination of your attributes may be constant across the alternatives
you show.**
:::

This is not exotic. It is the same phenomenon as the two normalizations everyone
already knows, fused into one condition:

| Classical model | Its classical problem | Standard fix |
|---|---|---|
| Multinomial logit | Adding a constant to all utilities changes nothing | Only differences are identified; no global intercept |
| Ordered logit | An intercept and the cut points cannot both move freely | Drop the intercept, or fix one cut point |
| **Our model** | **A constant in the design shifts $\overline{\mu}$ and trades off against the cut points** | **Reference- or effects-code; no global constant** |

: The dual-response condition is the union of the two classical ones {tbl-colwidths="[20,45,35]"}

Our model is a multinomial logit stacked on an ordered logit, so it inherits both
classical problems. The twist is that they are *linked*: the thing the ordered-logit
stage measures the level of is $\overline{\mu}_t(\beta)$, which is built from the same
$\beta$ the multinomial logit is estimating. That linkage is why the condition takes
the specific form it does.

## 1.3 The trap in practice: full dummy coding

The obvious violation is putting a literal intercept in the design. Nobody does that
on purpose. The violation that actually happens is **coding a categorical attribute
with all of its levels included**.

Suppose brand has three levels and you create three dummies: `brand_A`, `brand_B`,
`brand_C`. Every alternative is exactly one brand, so for every row of the design,

$$
\texttt{brand\_A} + \texttt{brand\_B} + \texttt{brand\_C} = 1 .
$$

There is your constant. Take $v = (1,1,1)$ across those three columns and you get
$\mathbf{x}_{tj}'v = 1$ for every good on every screen. You can add $1$ to all three
brand part-worths, add $1$ to every cut point, and nothing observable changes.

Here is that failure, exactly, using the paper's own likelihood function. The design
below has a constant column; we evaluate the negative log-likelihood at the true
parameters and at the shifted ones.

```{r d2}
des_bad <- make_design(n_tasks = 400, J = 3, seed = 1, intercept = TRUE)
id_rank_check(des_bad$X)   # the paper's rank condition: does it pass?
```

```{r d2b}
beta <- c(0.6, 1.0, -0.4, 0.5, -1.2, 0.0)  # last element multiplies the constant column
cut  <- c(-0.5, 0.7, 1.9)
dat  <- simulate_dual(des_bad, beta, cut, model = "B", seed = 7)

delta <- 1.3
par_true    <- c(beta, cut_to_par(cut))
par_shifted <- c(beta[1:5], beta[6] + delta, cut_to_par(cut + delta))

nll_true    <- negloglik(par_true, dat)
nll_shifted <- negloglik(par_shifted, dat)

cat(sprintf("nll at true parameters : %.12f\n", nll_true))
cat(sprintf("nll at shifted ones    : %.12f\n", nll_shifted))
cat(sprintf("difference             : %.3e\n", nll_shifted - nll_true))
cat(sprintf("bit-for-bit identical  : %s\n", identical(nll_true, nll_shifted)))
```

The difference is **exactly zero** — not small, not within tolerance, but bit-for-bit
identical. The likelihood is perfectly flat along that direction. Every point on that
ridge is equally good, so "the" maximum likelihood estimate does not exist, and which
answer you get depends entirely on where the optimizer happened to start.

The standard symptom is a Hessian with a zero eigenvalue:

```{r d3}
des_ok <- make_design(n_tasks = 400, J = 3, seed = 1, intercept = FALSE)
dat_ok <- simulate_dual(des_ok, beta[1:5], cut, model = "B", seed = 7)

c(`smallest eigenvalue, constant in design` = min(fit_dual_mle(dat)$eigen),
  `smallest eigenvalue, properly coded`     = min(fit_dual_mle(dat_ok)$eigen))
```

Numerically zero versus comfortably positive. This is the diagnostic to look at, and
it is why the paper reports the smallest eigenvalue of the observed information
alongside every simulation result.

**The fix is free.** Code categorical attributes against an omitted reference level
(two dummies for three brands), or effects-code them. One caveat worth stating: with
reference coding, the reference level has to actually *appear* in the design. If you
drop `brand_A` from the coding and then never show brand A, you have re-created the
constant.

## 1.4 Why the usual escape hatch is not available here

If you have run a standard choice model with a "none of these" option, you may be
thinking: *I include all my brand dummies all the time and it works fine.* You are
right, and the reason is instructive.

In the standard setup [@Haaijer_2001], the outside good sits on the screen as an
actual alternative with all-zero attributes. So when you shift the inside goods by
$\delta$, the outside good does **not** shift with them, the comparison changes, and
the probabilities move. The constant is identified because the row of zeros anchors
it.

```{r d4}
V <- c(1.2, -0.4, 0.8)

forced   <- rbind(original = softmax(V), shifted = softmax(V + 0.7))
with_out <- rbind(original = softmax(c(0, V)), shifted = softmax(c(0, V + 0.7)))

round(forced, 4)                     # our first response: no outside good on screen
round(with_out, 4)                   # standard model: outside good is a row of zeros
```

The top table does not move; the bottom one does. **In our design the outside good
never appears as a row.** It has been absorbed into the cut points — that is the whole
idea of the dual-response format, and at $W = 2$ the single cut point *is* the
no-choice constant of the standard model. Having given up that row, we have given up
the anchor that came with it, and anything the design holds constant across the inside
goods gets absorbed right along with it.

This is the one place where our identification requirement is genuinely stricter than
the familiar one, and it is worth flagging to anyone who brings habits from standard
CBC.

## 1.5 Normalizations versus restrictions

Two more conventions appear in the model, and it is worth being clear about which are
free and which are testable claims about the world.

**Free normalizations** (you must make some choice; nothing is being asserted):

- *The outside good has zero attributes.* This fixes the origin of the utility scale.
  Giving the outside good its own constant would shift all cut points equally, which is
  exactly the invariance above, so it is not a restriction we are imposing — it is one
  we could not avoid.
- *The shocks are standard Gumbel with scale 1.* This fixes the units of $\beta$, as in
  any logit. Doubling all utilities and all cut points changes nothing, so something
  has to set the scale.

**A genuine restriction** (testable, and we test it):

- *The coefficient on $\overline{\mu}_t$ in the second stage is exactly $1$.* This is
  not a normalization. It follows from the behavioral claim that the *same* utility
  draw produces both answers. You can free it up, estimate
  $\Pr(y_t \le w) = G(c_w - \lambda \overline{\mu}_t)$, and test $\lambda = 1$. That is
  an overidentification test of the model's central story, and the paper runs it.

The distinction matters for how you answer a referee. A normalization is a choice of
units and cannot be wrong. A restriction can be wrong, and the honest move is to test
it rather than to describe it as a normalization.

## 1.6 One more thing the design has to do

Proposition 1 guarantees identification of the cut points, but it does not guarantee they
are *well* estimated, and it is silent on a question we care about: **which link
function is right, Model A or Model B?**

Here is the issue. At a single value of $\overline{\mu}$, any two increasing link
functions can fit the observed category frequencies perfectly, just by relabeling the
cut points. There are $W-1$ frequencies and $W-1$ cut points; you can always solve it.
So a design in which every task is about equally attractive tells you *nothing* about
which link is correct.

What separates the links is how the answer distribution *moves* as the screen gets
better or worse. The two links agree in the middle and diverge in the tails, so you
need tasks out in the tails.

> **Practical implication.** If you want to distinguish Model A from Model B, or to
> estimate the cut points sharply, deliberately vary how attractive your screens are.
> A design optimized purely for the forced choice — which typically wants utility
> *balance*, all options roughly equally good — is close to the worst case for the
> second response.

This is a real tension between designing for question 1 and designing for question 2,
and it is treated at length in notebook 07.

# Section 2. The hierarchical model with common cut points

Now let respondents differ. Each has her own tastes,

$$
\beta_i \sim \mathcal{N}(\bar\beta, \Sigma),
$$

but everyone shares one set of cut points $\mathbf{c}$. The substantive assumption is
that people want different things but *use the words the same way*: "somewhat likely"
means the same thing to you as it does to me.

## 2.1 Three different questions, three different answers

The word "identified" now has to be applied to three separate things, and conflating
them is the main source of confusion in this literature.

| What you might ask about | Identified? | What does the work |
|---|---|---|
| The cut points $\mathbf{c}$ | Yes, easily | Pooling: every respondent's ordinal answers inform them |
| The population parameters $\bar\beta, \Sigma$ | Yes | Many respondents, $N \to \infty$ |
| One person's tastes $\beta_i$ | Weakly at best | Only her own handful of tasks |

: The three questions differ, and so do their answers {tbl-colwidths="[36,22,42]"}

Let us take them in order.

## 2.2 The cut points are the easy part

The cut points are *common*. So every ordinal answer from every respondent carries
information about them. With $N = 300$ respondents answering $T = 10$ tasks each, you
have $3{,}000$ ordinal answers informing $W - 1 = 4$ cut points.

That is an enormous amount of information per parameter. Cut points are estimated
precisely in this model, and they are essentially never the thing that goes wrong.

The same is true of $\bar\beta$ and $\Sigma$: they are features of the *population*, and
the population is what you sampled a lot of. Adding respondents drives their standard
errors down in the ordinary way.

## 2.3 Individual tastes are the hard part

Now the other direction. Respondent $i$ contributes $T$ choices and $T$ ordinal
answers — say $10$ and $10$ — and you want $P$ taste parameters from her, say $6$.

That is not much. Twenty observations for six parameters, where the observations are
categorical and each one is fairly uninformative. Her individual likelihood is a broad,
shallow hill rather than a sharp peak.

Worse, this does not improve as you collect more data in the usual sense. Adding
respondents adds *more* $\beta_i$ parameters, one per respondent, so you never
accumulate information about any particular one. Econometricians call this the
**incidental parameters** situation: the number of unknowns grows with the sample.

Here is what that looks like. We fit a single respondent's data on its own, with no
hierarchy, at three different task counts:

```{r d5}
small_design <- function(T_tasks, J = 3, seed) {
  set.seed(seed)
  n <- T_tasks * J
  list(X = cbind(hi = rbinom(n, 1, 0.5), price = runif(n, 0.5, 2.5)),
       n_tasks = T_tasks, J = J, P = 2)
}

sapply(c(8, 40, 200), function(Tt) {
  d  <- small_design(Tt, seed = 11)
  dd <- simulate_dual(d, beta = c(1.0, -0.8), cut = c(-0.4, 1.1),
                      model = "B", seed = 2)
  f  <- fit_dual_mle(dd)
  c(`se(beta_hi)` = f$se_beta[1], `se(c_1)` = f$se_cut[1])
}) |> `colnames<-`(paste0("T=", c(8, 40, 200))) |> round(3)
```

At $T = 8$ — a realistic commercial task count — the standard error on that
respondent's own taste parameter is larger than the parameter itself. On her own, she
tells you almost nothing.

## 2.4 What the hierarchy actually does, and what it does not

This is the point most worth getting right, because it is routinely overstated.

The hierarchical prior $\beta_i \sim \mathcal{N}(\bar\beta, \Sigma)$ says: respondents
are not arbitrary, they are draws from a population. When respondent $i$'s own data are
uninformative, her posterior is pulled toward what the rest of the sample looks like.
This is **shrinkage**, and it is genuinely valuable — it is why hierarchical Bayes
beats fitting each person separately, and it is why individual-level estimates from
CBC/HB are usable at all.

But be precise about what it is:

::: {.callout-warning}
## Shrinkage is borrowed information, not identification

Shrinkage improves your estimate of $\beta_i$ by importing information from other
respondents. That is legitimate — under the assumption that $i$ really is a draw from
the same population.

It does **not** repair a direction in which the model is truly unidentified. If a
direction is flat in *everyone's* likelihood, then borrowing from everyone borrows
nothing, and the posterior in that direction is just the prior, restated.

The distinction: shrinkage fixes *weak* identification (a shallow hill). It cannot fix
*non*-identification (a flat ridge).
:::

So which is it here? For the common-cut model, the good news is structural:

> **The common cut points act as an anchor for every individual.**

Because $\mathbf{c}$ is shared and pinned down by the whole sample, respondent $i$'s
*level* is measured against a scale that is already known. Even if her own design were
weak, the pooled data fix the ruler she is being measured with.

That means the rank condition from §1.2 has to hold on the **pooled** design, not on
each respondent's own. If your attribute coding is clean, it holds for everyone at once.
If you dummy-coded all levels, it fails for everyone at once — no rotation of the design
across respondents will save you, because the offending combination is constant in every
row of every screen shown to anybody.

**Summary for Section 2.** Nothing new is required beyond §1.2. Apply the same rank
condition to the pooled design and the model is identified. What changes is the
*precision* story: the population parameters and the cut points are sharp, individual
tastes are fuzzy and lean on the prior, and that is expected rather than alarming.

# Section 3. The hierarchical model with respondent-specific cut points

Now relax the shared-scale assumption. Each respondent gets her own cut points:

$$
c_{i1} \in \mathbb{R}, \qquad c_{iw} = c_{i,w-1} + e^{\delta_{iw}},
$$

with $\theta_i = (c_{i1}, \delta_{i2}, \ldots)$ drawn from a population distribution
jointly with $\beta_i$. (The exponentials just keep the cut points in increasing order
without constrained sampling. They are a computational device, not a modeling claim.)

This is a real phenomenon. Some people are enthusiasts who say "very likely" to
everything; others are reserved and never use the top box. @Greene_2010 and
@Falco_2015 document exactly this in other settings. But allowing it costs us something
specific, and this section is about what.

## 3.1 The new confound: enthusiasm versus generosity

Respondent $i$ answers "very likely" on nearly every task. Two explanations:

- **She likes everything.** Her $\beta_i$ is large; the products really do appeal to her.
- **She is a generous scale user.** Her $c_{i1}$ is low; she says "very likely" at a
  level of enthusiasm where you would say "somewhat."

From her ordinal answers *alone*, these are indistinguishable. High utility against
ordinary thresholds, and ordinary utility against low thresholds, produce identical
answers. This confound did not exist in Section 2, because there the thresholds were
common and known from everyone else.

## 3.2 What breaks the tie, and what does not

The confound is resolved — partly — by the forced choice.

Recall the division of labor from §1.1. The forced choice reveals the *contrast*
directions of $\beta_i$: whether she prefers red to blue, how she trades quality for
price. Those answers are completely unaffected by how she uses the words, because she
is not using the words when she picks. So the forced choice measures her tastes on a
scale that scale-use heterogeneity cannot touch, and given her tastes, the ordinal
answers reveal her thresholds.

**But only for the directions the forced choice can see.** And by §1.1, it cannot see a
direction that moves every alternative together. For that direction, we are back in
exactly the situation of §1.2 — with one important difference:

::: {.callout-important}
## The condition tightens: it must now hold respondent by respondent

In the **common-cut** model, a shift of $\beta_i$ had to be absorbed by the *shared*
cut points, and the rest of the sample would not stand for it. The pooled design was
what mattered.

In the **het-cut** model, respondent $i$'s shift can be absorbed by *her own*
$c_{i1}$. Nobody else's data constrain it, because nobody else's cut points moved.

So the rank condition must now hold on **each respondent's own design** — her $T \times J$
rows — not merely on the pooled design.
:::

In practice, commercial designs rotate attribute levels across respondents and include
at least one continuous attribute (price), so each respondent's own rows typically do
satisfy the condition. The paper says as much. But it is now a per-respondent
requirement that can fail for a subset of your sample, which is a thing worth checking
rather than assuming.

## 3.3 The counting problem gets materially worse

Set aside exact non-identification and just count. Per respondent, at $T = 10$ tasks,
$P = 6$ attributes, $W = 5$ scale points:

| | Parameters per respondent | Observations per respondent |
|---|:---:|:---:|
| Common cut points (§2) | $6$ ($\beta_i$ only) | $10$ choices $+$ $10$ ordinal answers |
| Respondent-specific cut points (§3) | $6 + 4 = 10$ | $10$ choices $+$ $10$ ordinal answers |

: Same data, many more individual parameters {tbl-colwidths="[40,30,30]"}

Ten parameters from ten ordinal answers and ten choices, where the four new parameters
are informed *only* by the ordinal half. Recall from §2.3 that at $T = 8$ the standard
error on a single cut point was already $1.6$ — on a utility scale where the whole
range of interest is a few units wide.

Individual cut points are therefore not really estimated from individual data. They are
almost entirely shrinkage. That is not fatal, but it does mean:

> Do not interpret one respondent's estimated cut points as a measurement of how *that
> person* uses the scale. What the model reliably recovers is the **population
> distribution** of scale use, and whether allowing it improves fit.

## 3.4 The specifically Bayesian hazard

Here is the part that deserves a warning label, and it applies to hierarchical Bayesian
models generally.

**A proper prior always produces a proper posterior.** If you put a normal prior on a
parameter and the likelihood is completely flat in some direction, the MCMC sampler will
run happily, the chains will mix, the traceplots will look fine, and you will get a
posterior with a mean and a credible interval. Nothing will crash.

But in that direction, the posterior *is the prior*. You have not learned anything; you
have printed your assumption back out with a standard error attached.

This is the essential difference between how the failure presents in the two estimation
methods:

| | Maximum likelihood (§1) | Hierarchical Bayes (§3) |
|---|---|---|
| Symptom | Optimizer fails; zero eigenvalue in the Hessian; different starts give different answers | Nothing. Chains mix, output looks normal |
| How you find out | Immediately, loudly | Only if you go looking |

: The same failure is loud in one framework and silent in the other {tbl-colwidths="[18,41,41]"}

Practical checks, in rough order of effort:

1. **Check the rank condition per respondent** before estimating anything. It is a
   `qr()` call on each respondent's design rows. This is cheap and catches the exact
   failure.
2. **Compare the posterior to the prior.** For any parameter where they coincide, the
   data said nothing. Widen the prior and see what moves; a genuinely identified
   parameter will barely notice.
3. **Watch $\beta_i$ and $c_{i1}$ together.** Near-unidentification in the level
   direction shows up as a strong negative correlation between a respondent's level-ish
   taste parameters and her first cut point, within her own posterior draws.

## 3.5 So should you use respondent-specific cut points?

The identification analysis gives a clear answer, and it is a conditional one.

Allowing heterogeneous cut points is the right call when you have reason to think scale
use genuinely varies and you have enough tasks per respondent to say something about it.
It is the wrong call when $T$ is small and your design was not built with per-respondent
rank in mind — there you will be reporting your prior.

The common-cut model is the safer default precisely because of the anchoring property in
§2.4: shared cut points pin down the ruler, so individual weakness is not fatal. The
paper's simulation study takes up the empirical side of this question — what you gain
when cut-point heterogeneity is real, and what you lose by ignoring it when it is not.

# A checklist

Everything above, compressed.

**Before you field the study**

- [ ] Reference-code or effects-code every categorical attribute. Never include all levels.
- [ ] Confirm the reference level actually appears in the design.
- [ ] Check that $[X \;\; \iota]$ has full column rank. One line of code; catches the fatal case.
- [ ] If you plan respondent-specific cut points, run that check *per respondent*, not just pooled.
- [ ] If you want to compare Model A against Model B, deliberately vary how attractive your screens are. Utility-balanced designs cannot distinguish the links.

**After you estimate**

- [ ] Report the smallest eigenvalue of the observed information (aggregate model).
- [ ] Check that every scale category was actually used; an empty category leaves its neighboring cut points weakly identified.
- [ ] Test $\lambda = 1$. It is a restriction, not a normalization, and it is the model's central behavioral claim.
- [ ] In the Bayesian models, compare posteriors to priors before believing individual-level output.

**Three sentences to remember**

1. The forced choice sees only differences; the ordinal report sees levels; anything constant across the alternatives is invisible to both.
2. Our design gave up the outside good's row of zeros, which is the anchor that lets standard choice models get away with full dummy coding — so we cannot.
3. Shrinkage fixes weak identification, not non-identification; in a Bayesian model the difference is invisible unless you check.

## References

::: {#refs}
:::
